Make any stupid plot you want

### 2021-02-11 draft/sections/intro.tex
Mackinlay + rework of visual channels section

This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:intro_dubois}; to do so, we first discuss the criteria by which a visualization is evaluated. As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design. He defines the graphical design as the set encoding relations from data to visual representation, and the design evaluated on an idealized space is what throughout this paper we will refer to as graphics \cite{mackinlayAutomatingDesignGraphical1986}. Mackinlay  proposes that a visualization tool can be evaluated by its expressiveness and effectiveness; A visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a}. Mackinlay's definition of expressiveness is formalized at the component level, for example that ordinal values must map to ordinal positions. He uses a scatter chart for discrete data and states that a line chart must have continuous data, but does not describe how to prove this. 
While Mackinlay is strictly focusing on the graphical components of a visualization, Byrne et al. propose that visualizations also have figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic nominal and numerical properties of the data are preserved in the graph. Tufte specifies that visual representations must be in proportion to the quantitative data being represented for a chart to be faithful and that there should be no extra information in the graphic or figurative elements of the graph, but otherwise his notion of graphic integrity is heavily context dependent\cite{tufteVisualDisplayQuantitative2001}. As is Norman's Naturalness Principal, which states that visualizations are more understandable  when the properties of the representation match the properties of the information being represented\cite{norman_things_smart}. Bertin takes it as a given that data properties match visual properties, so much so that Munzner argues it is inherently built into his classification system \cite{munznerVisualizationAnalysisDesign2014}. In this work, we propose that this property matching can be formally defined as equivariant maps from data to graphic representation.

\subsubsection{Retinal (Visual) Variables}
\label{sec:intro_visual_variables}
\begin{figure}[h!]
\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the \textit{components} of a visualization. This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} 
\cite{krygierMakingMapsVisual2005}}
\label{fig:retinal_variables}
\end{figure}
Figure~\ref{fig:retinal_variables} illustrates common guidelines for encoding \textit{components} of the data as components of the graphics. Bertin names these graphic components retinal variables; they are also called visual variables \cite{bertin_semiology_2011,krygier_making_2005,chambers_graphical_1983,wilkinson_grammar_2005,munzner_visualization_2014} or visual characteristics\cite{carpendaleVisualRepresentationSemiology}. The columns of figure~\ref{fig:retinal_variables} correspond to the type of observation: discrete points, continuous events (e.g. a timeseries), two dimensional continuous events (e.g. a vector map). The rows of figure~\ref{fig:retinal_variables} describe ways to visualize variations in the \textit{components}; generally, quantitative components are represented by retinal variables that change quantitatively and categorical components are represented by retinal variables that vary qualitatively. In figure~\ref{fig:iris_scatter}, the hue of the marker is used to encode differentiation in species and the position of the marker is used to show variation in petal length and sepal length. The retinal variables suggest that any single visualization is limited to encoding at most about 8 or 9 components. Retinal variables provide guidelines for encoding \textit{components}, but the choice of graph is based on the type and structure of the data.

>>>>

This work aims to develop a model of visualization such that a tool built using this model could support visualizations as varied as those of Du Bois in figure~\ref{fig:intro_dubois}; to do so, we first discuss the criteria by which a visualization is evaluated. Byrne et al. propose that visualizations have graphic representations that are mappings from data to visuals and figurative representations that have meaning due to their similarity in shape to external concepts \cite{byrneAcquiredCodesMeaning2016}. In figure~\ref{fig:intro_dbc}, Du Bois combines a graphical representation where glyph size varies by population with a figurative representation of those glyphs as the countries the data is from, which means that the semantic and numerical properties of the data are preserved in the graph. Tufte specifies that visual representations must be in proportion to the quantitative data being represented for a chart to be faithful and that there should be no extra information in the graphic or figurative elements of the graph, but otherwise his notion of graphic integrity is heavily context dependent\cite{tufteVisualDisplayQuantitative2001}. As is Norman's Naturalness Principal, which states that visualizations are more understandable  when the properties of the representation match the properties of the information being represented\cite{norman_things_smart}. Bertin takes it as a given that data properties match visual properties, so much so that Munzner argues it is inherently built into his classification system \cite{munznerVisualizationAnalysisDesign2014} which is displayed in figure~\ref{fig:intro_retinal_variables}.

\label{sec:intro_visual_variables}
\begin{figure}[H]
\includegraphics[width=1\textwidth]{figures/intro/retinal_variables.png}
\caption{Retinal variables are a codification of how position, size, shape, color and texture are used to illustrate variations in the components of a visualization. The best to show column describes which types of information can be expressed in the corresponding visual encoding. This tabular form of Bertin's retinal variables is from Understanding Graphics \cite{malamedInformationDisplayTips2010} who reproduced it from \textit{Making Maps: A Visual Guide to Map Design for GIS} 
\cite{krygierMakingMapsVisual2005}}
\label{fig:intro_retinal_variables}
\end{figure}

As described by Mackinlay, a visualization tool produces a graphical design and an image rendered based on that design. He defines the graphical design as the set encoding relations from data to visual representation\cite{mackinlayAutomatingDesignGraphical1986}, and the design rendered in an idealized abstract space is what throughout this paper we will refer to as a graphic. Mackinlay  proposes that a visualization tool's expressiveness is a measure of how much of the structure of the data the tool encodes, while the tools effectiveness describes how much design choices are made in deference to perceptual saliency \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014}. Mackinlay's definition of expressiveness is formalized at the visual encoding level, which as shown in figure~\ref{fig:intro_retinal_variables} refers to the components of a graphic such as the color or position of a glyph. Bertin first classified these graphic components as retinal variables and discussed which types of data they can express \cite{bertinIIPropertiesGraphic2011} and how they are composited on point, line, and area graphical marks, as shown in figure~\ref{fig:intro_retinal_variables} correspond. Mackinlay's expressiveness criteria is well defined for the visual variables, such that he suggests the viability of a strict encoding relation that is a homomorphic mapping which preserves some binary operator from one domain to another \cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987}. We expand on this suggestion by proposing that monoid action equivariance is a strict condition of building valid encoders.
